Sequential rank agreement methods for comparison of lists

### 2015-07-11 latex/paper.Rnw
smoothed abstract, intro, etc.

\begin{abstract}
  The comparison of rankings has received increased interest in the
  last decades. Computer science and genomics are just two areas that
  need to compare ranked lists in order to investigate similarities
  between different analytical methods. Often the aim is to find a
  measure of similarity for two or more ordered lists that comprise
  the first k items from each list. Since ordered lists typically are
  more similar at the top of the list another approach, which is taken
  in this paper is to ask: for how long are the lists highly similar?
  Using terminology from agreement analysis we derive the measure
  depth of agreement using the standard deviation of the ranks, a
  method that both provides an intuitive interpretation and can be
  applied to any number of lists even if these are incomplete. The
  finding are illustrated using simulations and applied to a publicly
  available cancer dataset.
\end{abstract}

Key words: sequential rank agreement, top lists, rank agreement, order statistic,
methods comparison, variable selection

@@

Ranking of items or results is common in scientific research and
ranked lists occur as the result of many applications in
statistics. Regression methods rank predictor variables according to
magnitude of association with outcome, prediction models rank subjects
according to their risk of an event, and genetic studies rank genes
according to their difference in gene expression levels across
samples.

When several ranked lists are available a common research question is
to what extent the lists agree on the ranking. In other words: is it
possible to decide on a optimal rank until which the lists agree on
the items?  A typical situation would arise in high-dimensional
genomics studies where several analysis methods had been applied to a
dataset or where a literature search provided differing tables of,
say, top 10 lists of the most important genes associated to an
outcome.  Similarly, this would correspond to the number of associated
predictor variables that are found consistently over the lists, the
number of patients at high risk of an event in all lists, and the
number of genes that are worth to pursue in further experiments in the
three examples above, respectively.

The analysis of rankings is hardly new, and several other approaches
have proposed a measure of the ``distance'' between two ranked
lists. Of these, Kendall's $\tau$
%\citep{kendall1948}
and Spearman’s footrule
%$\rho$ \citep{Spearman1910}
are the most well-known. Both measure the similarity of two ranked
lists but neither of them make any distinction between errors towards
the top versus the errors towards the bottom of the lists.  Several
modified versions of Kendall's $\tau$ and Spearman's footrule have been
proposed. \citet{Shieh1998} proposed a weighted $\tau$ where each pair
of rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
%
Spearman’s footrule use the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$
measure of \citet{Bar-Ilan2006} where the reciprocal rank differences
are used to calculate the similarity measure.

More recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersections also places
equal weights on all depths of the list which led \citet{Fagin2003}
and \citet{Webber2010} to modify the simple intersection to put more
emphasis on the top of the lists by using the cumulative intersection
divided by the depth as their similarity measures (denoted the average
overlap (AO)). \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series thus ensuring the top is
always weighted higher than the (potentially indefinite) bottom of the
lists.  While it is possible to calculate measures that describe how
similar lists are at a given depth with some of the methods described
above, the interpretation of the measures is unclear --- especially
with multiple lists --- and it is difficult to identify a depth at
which the lists are no longer similar.

In this article we propose a method to measure similarity among ranked
lists for each depth in the list. As a measure for similarity we will
use the limits of agreement known from agreement between quantitative
variables \citep{Carstensen2010} but will compute the agreement on the
ranks of the items.  The proposed method allows for multiple lists
simultaneously, provides a dynamic measure of agreement for each depth
in the lists, places more weight on the top of the list, accommodates
censored/incomplete lists of varying lengths, and has a natural
interpretation that directly relates to the ranks. The general idea is
to define agreement based on the sequence of ranks from the first $k$
elements in each list. This approach allows us to estimate a potential
change point in the agreement in order to determine an optimal $k$
where there is a shift in the agreement among lists and enables us to
use randomization approaches to evaluate the rank agreement. In this
sense it is a combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches.

The manuscript is organized as follows: In the next section we will
define the sequential rank agreement among multiple ranked lists and
discuss how to handle incomplete/censored lists. In section 3 we
present and discuss approaches to evaluate the sequential rank
agreement obtained from an experiment and to identify the optimal
depth $k$ for which the ranked lists agree. Finally we present three
quite different applications of the proposed sequential rank agreement
approach to real data before we discuss the findings along with
possible extensions.

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
and small values close to zero suggests that the lists agree on the
ordering while larger values suggests disagreement. If the ranked
lists are identical then the value of SRA will be zero for all depths
$d$.  The sequential rank agreement can be interpreted as the average
distance of the individual rankings of the lists from the average
ranking for each of the items we have seen until depth $d$. Note that
the terms that are part of \eqref{def:sra} are not independent since
each list contains the ranks from 1 tp $P$ exactly once. Hence we
simply use the pooled standard deviation as a measure of the average
distance between rankings.

\subsection{Agreement among fully observed lists}
The simplest case occurs when all $L$ lists are fully
observed, \ie, we have observed the rank of all $P$ items for all $L$
lists. Fully observed ranked lists are common and arise, for example,
when different statistical analysis methods are applied to a single
dataset to produce lists of predictors ranked according to their
importance, or if the same analysis method is applied to data from
different populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to improve the classification of acute leukemias
between two types: acute lymphoblastic leukemia (ALL) or acute myeloid
leukemia (AML) \citep{Golub1999}. Preprocessing of the gene expression
data was done as described in \citet{Dudoit2002} and the four
different analysis approaces were: marginal two-sample $t$ tests,
marginal logistic regression analyses, logistic regression eleastic
net, and marginal maximum information content correlations (MIC)
\citep{Reshef2011}. For the first two methods, the genes were ranked
according to $p$ value, for logistic regression the genes were ordered
by size of the corresponding coefficients (after standardization), and
MIC was ordered by correlation which resulted in the rankings seen in
Table~\ref{tab1}.  The sequential rank agreement at depth $d$ can be
interpreted as how many ranks apart do we on average expect the items
found until depth $d$ to be so smaller values on the $y$ axis
correspond to \emph{better} agreement among the lists than larger
values.

<<example1, echo=FALSE>>=
library(SuperRanker)
library(glmnet)
library(minerva)
library(changepoint)
library(randomForestSRC)

# Read data
library(multtest)
data(golub)

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE >>=
## Same as above but zoomed in
ysize <- 50
plot(res[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res2[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE >>=
## Same as above but zoomed in
ysize <- 50
plot(res2[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<label=tab1,echo=FALSE,results=tex>>=
library(xtable)
xxx <- as.data.frame(cbind(1:10, inputmatrix[1:10,]))
colnames(xxx) <- c("Ranking", colnames(xxx)[2:5])
xxx2 <- xtable(xxx, caption = "Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method")
digits(xxx2) <- 0
print(xxx2, table.placement = "tb", caption.placement = "top", include.rownames=FALSE)
#print(xtable(xxx,  label = "tab:one",

@@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig1a}
\includegraphics[width=.45\textwidth]{paper-fig2a}
\includegraphics[width=.45\textwidth]{paper-fig1b}
\includegraphics[width=.45\textwidth]{paper-fig2b}

\end{center}
 \caption{Left panels: Sequential rank agreement for 4 different analysis methods
   applied to the 3051 genes in the Golub data. Right panels:
   Corresponding sequential rank agreement for the same data but
   where only the top \Sexpr{topk} ranked items are available. The blue
   and red areas correspond to the independent and randomized
   reference hypothesis areas, respectively. The bottom plots are
   identical to the corresponding top plots but have been zoomed in on
   the first part of the $x$ axis.}
 \label{fig:example1}
\end{figure}

If the lists generally agree on all items then we expect the
sequential rank agreement curve to be flat, but if there are changes
in the levels of rank agreement then this suggests that there are sets
of items that the lists agree on and other sets where they rank the
items differently. In particular, we generally expect the agreement
among the lists to be better towards the top of the lists and worse
towards the end of the lists, and in this case the sequential rank
agreement curve will start at a low level and then increase until it
levels off.

\subsection{Analysis of incomplete/censored lists}
Incomplete lists are also a common occurrence that arise from, for
example, comparison of top $k$ lists, or when some methods only rank a
subset of the items (e.g., penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the predictors that have all been shrunk to zero).

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way.  Let $\Lambda_l, l=1, \ldots, L$ be the
set of items found in list $l$ so $\Lambda_l$ is the top $k_l$ list of
items from list $l$ where $k_l = |\Lambda_l|$. Note that if
$k_1=\cdots=k_L=k$ then we observe the top $k$ items for each of the
$L$ lists. For censored lists the rank function becomes
\begin{equation}
\tilde R_l(X_p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\
\{k_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some item is unknown in some of the lists.  The rankings within a
single list are clearly not independent since each rank must appear
exactly once in each list. Thus we cannot simply assign the same
number (e.g., the mean of the unassigned ranks) to the censored items
since that would result in less variation of the agreement, and it
would artificially introduce a (downward) bias of agreement for items
that are censored in multiple lists.

Instead we assume that the rank assigned to an item $X_p$ that is not
present in list $l$ (i.e., is censored) is uniformly distributed among
the ranks that have \emph{not} been assigned for list $l$. By
randomizing unobserved items we attain one realization of the rankings
where all lists are fully observed. By randomizing a large number of
times we can compute \eqref{def:sra} for each realization, and then
compute the sequential rank agreement as the pointwise (for each
depth) average of the rank agreements. The algorithm is described in
detail in Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
\Procedure{Censored rank agreement}{}
\State Let $B$ be the number of permutations to use
\For{each $b \in B$}
\For{each censored list $l \in L$}
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute the unassigned ranks, $k_l+1, \ldots, P$ and assign
them randomly to the items \emph{not} found in the list, i.e.,
$\Lambda^\complement$, in order to fill out the lists.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the items
of interest will be found towards the top of the lists, and 2) that
the censored rankings provide so little information that it is
reasonable to assume that the missing rankings provide no information
and might as well be considered random.  The first assumption is
justifiable because we have already accepted that it is reasonable to
rank the items in the first place. The second assumption is fair
in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are combined then it is clear that the
interesting part of the sequential rank agreement curves must be for
smaller depths. Unless we have additional prior knowledge about the
distribution of the unranked items then we can restrict attention of
the sequential rank agreement to
\begin{equation}
d' \leq \max(k_1, \ldots, k_L),
\end{equation}
since we have no real information about the ordering of the items after a
depth $d'$ where all lists are censored.

@@

\section{Evaluating sequential rank agreement}

To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists}
\end{eqnarray*}
which not only assumes that there is no information in the rankings
but also that the methods are completely independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
H_1 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is clearly the strictest (and also potentially quite
unrealistic) null hypothesis but we can easily obtain realizations
from that null hypothesis simply by permuting the items for each list
and then computing the sequential rank agreement for the permuted
lists. In the uncensored case each experiment contains $L$ lists of
random permutations of the numbers $1,\dots,P$. For the censored case
we permute the number as before but censor the same number of items
for each list as were censored in the original dataset. The sequential
rank agreement curve from the the original data can then be compared
to the pointwise rank agreements obtained from under $H_0$.

To obtain a reference distribution under $H_1$ we use a randomization
test approach on the original data to remove any associations to the
outcome in the data that was used to generate the rankings before
computing the ranking and subsequently the sequential rank agreement
from the randomized data. The idea is to repeat the ranking procedures
in the modified data where the outcome vector is randomly permuted. In
practice we permute the outcomes but keep any predictors before
ranking to preserve any structure in the predictors. This artificially
removes any potential information regarding the regression that the
real data may contain. The randomization approach requires that we
have the original data available, so we can compute the ranked lists
from the randomized data and not just the ranked lists and as such it
may not be possible to evaluate $H_1$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $H_1$ then this suggests that the
original ranked lists agree \emph{more} that we would expect if the
ranked lists were based on data with no information.

Figure~\ref{fig:example1} shows the sequential rank distributions
based on $400$ samples under $H_0$ and $H_1$, respectively. Not
surprisingly, the null distribution of $H_1$ has lower sequential rank
agreement than the null distribution of $H_0$ because the four methods
used to rank the data ($t$ test, logistic regression, elastic net, and
MIC) generally identify and rank the same set of predictors that show
spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement is better than what would be expected for the full
dataset but that the censored data (the bottom-right plot) suggests
that there may be at most 1 or 2 ranked items that are towards the top
of the lists that yield a result better than what would be expected.

It is important to stress that neither $H_0$ nor $H_1$ are concerned
with whether the $L$ rankings are relevant. Both hypotheses are purely
considering how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

@@

The average overlap is widely used in comparing (two) ranked list and
the idea behind the average overlap closely resembles the sequential
rank agreement \citep{Fagin2003,Webber2010}. Here we will present an
extention of the average overlap to more that two lists and compare this
to sequential rank agreement.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  indentify change points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

Key words: sequential rank agreement, order statistic, gene rankings,
methods comparison, variable selection, permutation

@@

A common task of scientific research is to rank items obtained by some
form of data analysis. Ranked lists occur naturally in applications of
statistics. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent do they agree. Once agreement is
established a further question is regarding the depths of agreement
starting from the top of the ranked lists. A typical situation arises
in high-dimensional genomics studies when several analysis methods are
applied to rank a list of genes according to their association with a
phenotype, treatment effect or other outcome. A measure of agreement
of gene rankings obtained by different methods could often help to
identify which genes are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known. But these measures do not distinguish between agreement in
the top versus towards the bottom of the lists. \citet{Shieh1998}
proposed a weighted version of Kendall's $\tau$ where each pair of
rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
% 
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the agreement measure.

More recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list. Thus, \citet{Fagin2003} and
\citet{Webber2010} proposed modified intersections which put more
emphasis on the top of the lists. One proposal is denoted the average
overlap (AO) and used for comparison later on. Specifically,
\citet{Webber2010} define their rank-biased overlap (RBO) by weighting
with a converging series to ensure that the top is weighted higher
than the potentially non-informative bottom of the lists.

It is possible to use the existing methods to calculate agreement of
lists until a given depth, i.e., limited to the $d$ items of each
list. However, the interpretation may not be straightforward,
especially in the case of more than two lists.

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists. We provide graphical tools to analyse
the depth of agreement of a set of ranked lists. The general idea is
to define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{Carstensen2010}. Our proposed approach allows to compare more
than two lists simultaneously, provides a dynamic measure of agreement
as a function of the depth in the lists, places high weight on the top
of the list, accommodates censored/incomplete lists of varying
lengths, and has a natural interpretation that directly relates to the
ranks. The graphical illustration of sequential rank agreement
potentially allows to infer a changepoint, i.e., a list depth where
there is a change in the agreement of the lists.

In settings where rankings are obtained from analysis of correlated
variables it is then possible to compare the actual results to the
expected agreement in non-informative data obtained by
randomization. In this sense our approach is a combination and
generalization of some of the ideas of \citet{Carterette2009} and
\citet{Boulesteix2009}. The former compares two rankings based on the
distance between them as measured by a multivariate Gaussian
distribution and the latter presents an overview of approaches for
aggregation of ranked lists including bootstrap and leave-one-out
jackknife approaches.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored lists. In section 3 we present and discuss
approaches to evaluate the sequential rank agreement obtained from an
experiment and to identify the maximal depth until which the ranked
lists agree to certain extent. Finally we present three different
applications of the proposed sequential rank agreement approach to
real data before we discuss the findings along with possible
extensions.

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}. \label{def:sra}
\end{equation}
Values of \textrm{SRA} close to zero suggests that the lists agree
while larger values suggests disagreement. If the ranked lists are
identical then the value of SRA will be zero for all depths $d$. The
sequential rank agreement can be interpreted as the average distance
of the individual rankings of the lists from the average ranking for
each of the items we have seen until depth $d$. Note that the terms
that are part of \eqref{def:sra} are not independent since each list
contains the ranks from 1 to $P$ exactly once. Hence we simply use the
pooled standard deviation as a measure of the average distance between
rankings.

\subsection{Agreement among fully observed lists}
\label{sec:amfol}

The simplest case occurs when all $L$ lists are fully observed, \ie,
we have observed the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaces were: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the rankings seen in Table~\ref{tab1}.
The sequential rank agreement shown in Figure~\ref{fig:example1} shows
for each depth $d$ how many ranks apart do we on average expect the
genes found in the top $d$ part of the lists. It can be seen that
agreement is higher in the top of the lists (smaller values on the $y$
axis) correspond to \emph{better} than in the tails of the lists.

<<example1, echo=FALSE, eval=FALSE>>=
library(SuperRanker)
library(glmnet)
library(minerva)
library(changepoint)
library(randomForestSRC)
# Read data
library(multtest)
data(golub)

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=FALSE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
## Same as above but zoomed in
ysize <- 50
plot(res[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res2[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE, eval=FALSE>>=
## Same as above but zoomed in
ysize <- 50
plot(res2[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

<<label=tab1,echo=FALSE,results=tex, eval=FALSE>>=
library(xtable)
xxx <- as.data.frame(cbind(1:10, inputmatrix[1:10,]))
colnames(xxx) <- c("Ranking", colnames(xxx)[2:5])
xxx2 <- xtable(xxx, caption = "Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method")
digits(xxx2) <- 0
print(xxx2, table.placement = "tb", caption.placement = "top", include.rownames=FALSE)
#print(xtable(xxx,  label = "tab:one",

@@

\begin{figure}[tb]
   \begin{center}
 \includegraphics[width=.45\textwidth]{paper-fig1a}
 \includegraphics[width=.45\textwidth]{paper-fig2a}
 \includegraphics[width=.45\textwidth]{paper-fig1b}
 \includegraphics[width=.45\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data. Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

If the lists would agree on all items then we would expect the
sequential rank agreement curve to be flat. And if there are changes
in the levels of rank agreement then this would suggest that there are
sets of items for which the lists are in high agreement and sets of
other items where the lists rank the items differently. In gene
association studies we generally expect the agreement among the lists
to be better towards the top of the lists and worse towards the end of
the lists. In this case the sequential rank agreement curve will start
at a low level and then increase until it levels off as in
Figure~\ref{fig:example1}.

\subsection{Analysis of incomplete/censored lists}

Incomplete lists are also a common occurrence. They arises for example
in case of missing data (items) or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero. Incomplete lists also occur
when the analyst decides to censor all lists at depth $d$ such that
the measurement of agreement is restricted to the first $d$ elements
of each list.

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \tilde R_l(X_p) = \left\{\begin{array}{cl} \{\pi_l^{-1}(p)\} & \text{ for } p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, the rankings within a single
list are not independent since each rank must appear exactly once in
each list. Thus, we cannot simply assign the same number (e.g., the
mean of the unassigned ranks) to the censored items since that would
% @Claus: less variation of the ranks? or of the agreement?
result in less variation of the agreement, and it would artificially
introduce a (downward) bias of agreement for items that are censored
in multiple lists.

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. By randomizing the missing items of
each list once we attain one realization of the $L$ rankings of the
set $\Lambda_P$. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Censored rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each censored list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=\Lambda_P\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the most
interesting items are found in the top of the lists, and 2) that the
censored rankings provide so little information that it is reasonable
to assume that they can be represented by a random order. The first
assumption is justifiable because we have already accepted that it is
reasonable to rank the items in the first place. The second assumption
is fair in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the percentage of censored items is
low. Generally, without additional prior knowledge about the
distribution of the censored items it seems reasonable to restrict the
attention of the sequential rank agreement to depths smaller than
$\max(d_1, \ldots, d_L)$.

@@

\section{No information agreement}

We now consider the special case where the rankings are obtained by
applying different methods to the same data set which contains one
outcome variable and a list of candidate predictors. To interpret the
sequential rank agreement values in this situation we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

The first reference hypothesis is
\begin{eqnarray*}
H_0  & : &  \text{The list rankings correspond to complete randomly}\\
       &  & \text{permuted lists}.
\end{eqnarray*}
This does not only assume that there is no information data on which
in the rankings are based but also that the methods are completely
independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
% @Claus: maybe H_1 is not a good name as people would expect H_1 to be the complement of H_0
\begin{eqnarray*}
\tilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

To obtain the distribution under $\tilde H_0$ the idea is to repeat
many times the ranking procedures in uncorrelated data.  Thus, we
first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\tilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\tilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information.

For the analysis of the Golub data described in Section \ref{sec:amfol},
Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\tilde H_0$ each based on
$400$ permutations. Not surprisingly, the sequential rank agreement
under $\tilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank high) similar predictors even if there are only spurious
associations. The two bottom panels in
% @Claus jeg forstaar ikke den den naste lange saetning. mener du full lists i stedet for full dataset?
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement is better than what would be expected for the full
dataset but that the censored data (the bottom-right plot) suggests
that there may be at most 1 or 2 ranked items that are towards the top
of the lists that yield a result better than what would be expected.

It is important to stress that neither $H_0$ nor $\tilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the data set. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

@@

The average overlap is widely used in comparisons of (two) ranked
lists \citep{Fagin2003,Webber2010} and the idea behind the average
overlap closely resembles the sequential rank agreement. In this
section we present an extention of the average overlap to more that
two lists and compare this to sequential rank agreement.
